Spatial variation in predator diel activity patterns -- feral cats avoid red foxes in time, not space

### 2022-04-15 manuscript/diel-manuscript-supp.Rmd
submission1-JAE

---
title: "Spatial variation in predator diel activity patterns -- feral cats avoid red foxes in time, not space"
author:
  - name: Matthew W. Rees
    email: matt.wayne.rees@gmail.com
    affiliation: UOM
    footnote: 1
  - name: Brendan A. Wintle
    email: 
    affiliation: UOM
  - name: Jack H. Pascoe
    email: 
    affiliation: CEC
  - name: Mark Le Pla
    email: 
    affiliation: CEC
  - name: Emma K. Birnbaum
    email: 
    affiliation: CEC
  - name: Bronwyn A. Hradsky
    email: 
    affiliation: UOM
    footnote: 
address:
  - code: UOM
    address: Quantitative & Applied Ecology Group, School of Ecosystem and Forest Science, The University of Melbourne, Parkville, VIC, Australia
  - code: CEC 
    address: Conservation Ecology Centre, Otway Lighthouse Rd, Cape Otway, VIC, Australia
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
keywords: "feral cat; diel activity patterns; generalised additive model; invasive predator;  intraguild predator interactions; mesopredator release; red fox; lethal predator control"
journal: "?"
date: "`r Sys.Date()`"
bibliography: /Users/mrees2/Dropbox/personal/matt/github/phd-thesis/index/bib/thesis.bib
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{float}
    - \floatplacement{figure}{H}
    - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}} \setcounter{section}{0} \renewcommand{\thesection}{S\arabic{section}}}
    - \usepackage{lineno}
linenumbers: false
csl: journal-of-applied-ecology.csl
geometry: margin=1.1in
fontsize: 12pt
output:
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
editor_options: 
  chunk_output_type: inline
---

@@

*Article type*: Research Article

*Running title*: Spatiotemporal predator activity

*Word count (abstract, body)*: 330, 4338

*References*: ?

*Figures*: 5

*Tables*: 0

*Authorship*:

*Data accessibility*:   Data and code will be deposited on the Dryad Digital Repository after acceptance and can be viewed here: [https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators](https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators).

@@

Understanding the constraints that apex predators impose on subordinate species is important for anticipating the outcomes of predator management. Subordinate predators may avoid dominant predators in time or space, making it difficult to quantify changes in antipredator behaviours unless joint spatiotemporal analyses are used.

In this study, we tested whether an invasive apex predator (red fox *Vulpes vulpes*) suppresses or alters the spatiotemporal activity of an invasive mesopredator (feral cat *Felis catus*). We surveyed these predators using 3667 camera-trap deployments across two regions of south-eastern Australia; foxes were poison-baited in some landscapes within each region. The simple predator guild in these regions allowed sharp focus on the interactions between these species across experimental gradients of fox (apex predator) activity. We used generalised additive models to quantify overall predator activity across space and fluctuations in predator activity throughout the daily cycle (i.e., diel activity patterns).

When averaged across the study region, red foxes and feral cats had very similar diel activity patterns; however, there was important differentiation at a finer scale. When fox counts at a camera-trap were high, feral cats did not reduce their overall activity but shifted their diel activity patterns to less risky times of the day. In dry habitats of both regions, cats shifted from being nocturnal-crepuscular to mostly diurnal. In wet forest habitat, fox activity was consistent throughout the diel period; but when fox counts were high, cats became more nocturnal, avoiding dawn in particular. Changes in cat diel activity patterns may facilitate spatial coexistence between the two invasive predators, potentially shifting impacts onto different native prey species.

It is well-appreciated that overall predator activity varies spatially and fluctuates throughout the daily cycle. Our study demonstrates that diel activity patterns also vary across space, likely mediated by both landscape context and fear. Apex predator avoidance appears to be dynamic across landscapes of fear—a  key nuance which is overlooked when simply comparing the average activity overlap between two species or the spatial overlap of species occurrence.

@@

The red fox *Vulpes vulpes* (hereafter 'fox'; ~6 kg) and feral cat *Felis catus* (hereafter 'cat'; ~4 kg) have devastating impacts on native prey throughout their introduced range, implicated in the extinction of ~10 and 63 species, respectively [@doherty2016invasive]. The impacts of these invasive predators have been particularly extreme on the Australian continent [@woinarski2015ongoing]. Cats are more difficult to manage, and so introduced predator control programs in Australia often target only foxes [particularly through poison-baiting; @reddiex2007control]. As foxes and cats compete for many of the same resources, there is concern that lethal fox control could cause a mesopredator release [@soule1988reconstructed] of feral cats [@glen2005complex; @doherty2017stop; @wayne2017recoveries; @marlow2015cats; @comer2020integrating; @robley2014long]. There is some evidence that feral cats increase in activity [although highly uncertain; @hunter2018not], density (Chapter \@ref(density)) and alter their diets and use of space [Chapter \@ref(density); @molsher2017mesopredator] in response to fox control. Other studies have investigated potential spatial and temporal interactions between these invasive predators [e.g., @roshier2021space], but not in response to fox control, or in a joint spatiotemporal framework that allows flexibility in cat avoidance behaviours in respect to differences in fox activity.

@@

I had field assistance from: Shauni Omond, Shayne Neal, Asitha Samarawickrama, Shelley Thompson, Erin Harris, Hannah Killian, Lani Watson, Mark Dorman, Jack Davis, Carl Roffey, Bruce Edley, Larissa Oliveira Gonçalves, Ben Lake, Chantelle Geissler, Aviya Naccarella, Emily Gronow, Harley England, David Pitts, Annie Hobby, Louise Falls, Thomas McKinnon, Jimmy Downie, Marney Hradsky, Stephanie Samson, Robin Sinclair, Asmaa Alhusainan, Kelly Forrester, Tammana Wadawani, Emily Gregg, Hannah Edwards, Adam Beck, Vishnu Memnon, Sandy Lu, Pia Lentini, Nick Golding, Emily McColl-Gausden, Nina Page, Maggie Campbell-Jones, Kyle Quinn and Jack Dickson.

Surveys were conducted under University of Melbourne Animal Ethics Committee approval 1714119 and Victorian Government Department of Environment, Land Water and Planning Research Permit 10008273.

@@

\beginsupplement

# Supporting Information {-}

>>>>

---
title: "Spatial variation in predator diel activity patterns -- feral cats avoid red foxes in time, not space"
author:
  - name: Matthew W. Rees
    email: matt.wayne.rees@gmail.com
    affiliation: UOM
    footnote: 1
  - name: Brendan A. Wintle
    email: 
    affiliation: UOM
  - name: Jack H. Pascoe
    email: 
    affiliation: CEC
  - name: Mark Le Pla
    email: 
    affiliation: CEC
  - name: Emma K. Birnbaum
    email: 
    affiliation: CEC
  - name: Bronwyn A. Hradsky
    email: 
    affiliation: UOM
    footnote: 
address:
  - code: UOM
    address: Quantitative & Applied Ecology Group, School of Ecosystem and Forest Science, The University of Melbourne, Parkville, VIC, Australia
  - code: CEC 
    address: Conservation Ecology Centre, Otway Lighthouse Rd, Cape Otway, VIC, Australia
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
keywords: "feral cat; diel activity patterns; generalised additive model; invasive predator;  intraguild predator interactions; mesopredator release; red fox; lethal predator control"
journal: "Journal of Animal Ecology"
date: "`r Sys.Date()`"
bibliography: /Users/mrees2/Dropbox/personal/matt/github/phd-thesis/index/bib/thesis.bib
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{float}
    - \floatplacement{figure}{H}
    - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}} \setcounter{section}{0} \renewcommand{\thesection}{S\arabic{section}}}
    - \usepackage{lineno}
linenumbers: false
csl: journal-of-animal-ecology.csl
geometry: margin=1.1in
fontsize: 12pt
output:
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
editor_options: 
  chunk_output_type: inline
---

@@

@@

1. Understanding the constraints that apex predators impose on subordinate species is important for anticipating the outcomes of predator management. Subordinate predators may avoid dominant predators in time or space, making it difficult to quantify changes in antipredator behaviours unless joint spatiotemporal analyses are used.

2. In this study, we tested whether an invasive apex predator (red fox *Vulpes vulpes*) suppresses or alters the spatiotemporal activity of an invasive mesopredator (feral cat *Felis catus*).

3. We surveyed these predators using 3667 camera-trap deployments across two regions of south-eastern Australia; foxes were poison-baited in some landscapes within each region. The simple predator guild in these regions allowed sharp focus on the interactions between these species across experimental gradients of fox (apex predator) activity. We used generalised additive models to quantify overall predator activity across space and fluctuations in predator activity throughout the daily cycle (i.e., diel activity patterns).

4. When averaged across the study region, red foxes and feral cats had very similar diel activity patterns; however, there was important differentiation at a finer scale. When fox counts at a camera-trap were high, feral cats did not reduce their overall activity but shifted their diel activity patterns to less risky times of the day. In dry habitats of both regions, cats shifted from being nocturnal-crepuscular to mostly diurnal. In wet forest habitat, fox activity was consistent throughout the diel period; but when fox counts were high, cats became more nocturnal, avoiding dawn in particular. Changes in cat diel activity patterns may facilitate spatial coexistence between the two invasive predators, potentially shifting impacts onto different native prey species.

5. It is well-appreciated that overall predator activity varies spatially and fluctuates throughout the daily cycle. Our study demonstrates that diel activity patterns also vary across space, likely mediated by both landscape context and fear. Apex predator avoidance appears to be dynamic across landscapes of fear&mdash;a key nuance which is overlooked when simply comparing the average activity overlap between two species or the spatial overlap of species occurrence.

@@

The red fox *Vulpes vulpes* (hereafter 'fox'; ~6 kg) and feral cat *Felis catus* (hereafter 'cat'; ~4 kg) have devastating impacts on native prey throughout their introduced range, implicated in the extinction of ~10 and 63 species, respectively [@doherty2016invasive]. The impacts of these invasive predators have been particularly extreme on the Australian continent [@woinarski2015ongoing]. Cats are more difficult to manage, and so introduced predator control programs in Australia often target only foxes [particularly through poison-baiting; @reddiex2007control]. As foxes and cats compete for many of the same resources, there is concern that lethal fox control could cause a mesopredator release [@soule1988reconstructed] of feral cats [@glen2005complex; @doherty2017stop; @wayne2017recoveries; @marlow2015cats; @comer2020integrating; @robley2014long]. There is some evidence that feral cats increase in activity [although highly uncertain; @hunter2018not], density (M.W.R, unpublished data) and alter their diets and use of space [@molsher2017mesopredator] in response to fox control. Other studies have investigated potential spatial and temporal interactions between these invasive predators [e.g., @roshier2021space], but not in response to fox control, or in a joint spatiotemporal framework that allows flexibility in cat avoidance behaviours in respect to differences in fox activity.

@@

We acknowledge the Gunditjmara and Eastern Maar peoples on whose traditional lands this study took place. We thank Nick Golding, who inspired the use of GAMs in this paper and provided statistical advice. We are grateful to Alan Robley, the Department of Environment, Land, Water, and Planning (DELWP) Heywood team and Parks Victoria West Coast District Team for sharing camera-trapping datasets. The M.W.R PhD surveys were conducted under University of Melbourne Animal Ethics Committee approval 1714119.5 and Victorian Government DELWP Research Permit 10008273, with the assistance of Shauni Omond, Shayne Neal, Asitha Samarawickrama, Shelley Thompson, Erin Harris, Hannah Killian, Lani Watson, Mark Dorman, Jack Davis, Carl Roffey, Bruce Edley, Larissa Oliveira Gonçalves, Ben Lake, Chantelle Geissler, Aviya Naccarella, Emily Gronow, Harley England, David Pitts, Annie Hobby, Louise Falls, Thomas McKinnon, Jimmy Downie, Marney Hradsky, Stephanie Samson, Robin Sinclair, Asmaa Alhusainan, Kelly Forrester, Tammana Wadawani, Emily Gregg, Hannah Edwards, Adam Beck, Vishnu Memnon, Sandy Lu, Pia Lentini, Nick Golding, Emily McColl-Gausden, Nina Page, Maggie Campbell-Jones, Kyle Quinn and Jack Dickson. This study was generously supported by the Conservation Ecology Centre, the Victorian Government DELWP, Parks Victoria, Holsworth Wildlife Research Endowment – Equity Trustees Charitable Foundation and Ecological Society of Australia, Australian Research Council Linkage Project LP170101134, and the Australian Government’s National Environmental Science Program through the Threatened Species Recovery Hub. M.W.R also received funding from an Australian Government Research Training Program Scholarship and a Victorian Government DELWP top-up scholarship.

CONFLICT OF INTEREST {-}
===================
The authors declare no conflicts of interest.

AUTHOR CONTRIBUTIONS {-}
===================
M.W.R conceived the ideas and designed the methodology; M.W.R, J.H.P, M.LP, E.K.B and B.A.H collected some of the data; M.W.R analysed the data with input from B.A.H and B.A.W, and led the writing of the manuscript. All authors contributed critically to the drafts and gave final approval for publication.

DATA AVAILABILITY {-}
===================
Data and code will be deposited on the Dryad Digital Repository after acceptance and can be viewed here: [https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators](https://github.com/matt-w-rees/spatiotemporal-gams-invasive-predators).

@@

\newpage
\beginsupplement
\setcounter{page}{1}
\nolinenumbers

# SUPPORTING INFORMATION {-}

*Spatial variation in predator diel activity patterns -- feral cats avoid red foxes in time, not space*

MW Rees, BA Wintle, JH Pascoe, M Le Pla, EK Birnbaum & BA. Hradsky
